---
title: "Practical Machine Learning Final Assignment"
author: "Simon Grasdal"
date: "2024-12-23"
output:
  pdf_document: default
  html_document: default
---

```{r Setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(e1071)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
```

# Background
One thing that people regularly do is quantify how  much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

# Introduction
Because of a group of movement enthusiasts, the quantified self movement have recorded an astounding amount of data to determine how often movement are performed. Using devices from brands like Jawbone Up, Nike FuelBand, and Fitbit this data can be precisely measured and stored. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

In this assignment, we use accelerometer data from the wait (belt), forearm, are and dumbbells of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. Using a machine learnign algorithm, we attempt to predict the *manner* in which each participant doe the excercise. 

# Data Cleaning/Preparation
### Importing the Data
``` {r Importing the Data}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
```

### Reading and CLeaning the Data
```{r Reading and Cleaning the Data}
trainRaw <- read.csv("./data/pml-training.csv")
testRaw <- read.csv("./data/pml-testing.csv")
dim(trainRaw)
dim(testRaw)

# Identifying Complete Cases
sum(complete.cases(trainRaw))

# Removing Na's
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 

# Removing Columns that do not contribute to performace
classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]
```
The training set rendered has 19,220 observations and 53 total variables. The Testing set contains 20 observations and likewise 53 variables. 

# Slicing the Data
The cleaned training set is split into a validation data set (30%) and a pure training data set (70%). The validation data set assists in conducting cross validation.
``` {r Slicing the Data}
set.seed(20420) # For reproducible purpose
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]

```

# Modeling the Predictive Algorithm 
Using a **Random Forest** algorithm we fit a predictive model for activity recognition. This algorithm was chosen because it automatically selects key variables and is robust. The *5-fold cross-validation* is used when applying the algorithm.
```{r Modeling}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```

``` {r}
predictRf <- predict(modelRf, newdata = testData)
#Testing accuracy
confusionMatrix(table(predictRf, testData$classe))
```

The quality of model fit for the prediction can be determined by calculating the values of accuracy and the out-of-sample Root Mean Square Error (RSME). In the interests of cross-validation, the RSME was normalized to aid in interpreting how well the prediction model fitted the test data.

The estimated accuracy of the model is 99.79% and the Normalized out-of-sample error (RMSE) has a low value between 1 and 0  (0.00815633). These results indicate a high degree of fit for the prediction of the dataset.

``` {r}
accuracy <- postResample(table(predictRf), table(testData$classe))
accuracy
oose <- 1 - as.numeric(confusionMatrix(table(testData$classe, predictRf))$overall[1])
oose
```

# Final Model (Top 20 Predictor Variables)
``` {r}
modelRf$finalModel
varImp(modelRf)
```


# Conclusion
A machine learning (ML) model predicted the manner of participant exercise, which was Classe ‘A’. Accelerometer data located on participants’ belt, forearm, arm, and dumbell, from an Exercise dataset, was cleaned, and split into training and test datasets. The predictive model successfully identified the Classe. RSME calculations contributed to the anticipated strength of the model developped for this data.


# Appendix
##Figures - Data Visualization

### Figure 1. - Correlation Matrix Visualization
```{r}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```

### Figure 2. - Decision Tree Visualization
``` {r}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) # fast plot
```

### Figure 3. - Prediction Model 
The prediction model was applied to the original testing data set, downloaded from the data source.
``` {r}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```









